---
title: Canadian Stock Market and Inflation
author: JY
date: '2022-02-27'
slug: canadian-stock-market-and-inflation
categories:
  - R
tags:
  - ''
description: ''
thumbnail: ''
---

Exploration of relationship between inflation and stock market returns in Canada.

```{r setup, includes=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warning=FALSE,
                      fig.height=4,
                      fig.width=8)
options(scipen = 99)
options(digits=3)

library(tidyverse)
library(here)
library(lubridate)
```

### Get Data

Data sourced from Statistics Canada:

* Canadian inflation data: [source]
* Toronto Stock Exchange data: https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1010012501

```{r}
infl_data <- read_csv(here('data','statscan-inflation-smry-1970.csv'))
infl_data <- infl_data %>% rename(
  date=REF_DATE,
  prod_groups=`Products and product groups`
)
tse_data <- read_csv(here('data','tse-1970-relevant-stat.csv'))
```

Check out data:

* inflation date ranges & categories

```{r}
summary(infl_data$date)
infl_data %>% count(prod_groups)
```

* TSE date ranges & categories

```{r}
summary(tse_data$date)
tse_data %>% count(TSE_stat)
```

Start with highest-level categories

```{r}
infl_data_main <- infl_data %>% filter(prod_groups=='All-items')
infl_data_comb <- infl_data_main %>% select(-GEO, -COORDINATE)
infl_data_comb <- infl_data_comb %>% rename(
  item=prod_groups
)

tse_data_main <- tse_data %>% filter(TSE_stat=='S&P/TSE Composite Index, close')
tse_data_comb <- tse_data_main %>% select(-COORDINATE, -UOM_ID, -TERMINATED)
tse_data_comb <- tse_data_comb %>% rename(
  item=TSE_stat
) %>% mutate(
  item=str_replace_all(item, 'S&P/TSE Composite Index, close','TSE Index close')
)

```

Calc % change for comparison

```{r}
infl_data_comb_pc <- infl_data_comb %>% mutate(
  pc_chg=VALUE/lag(VALUE)-1
)
tse_data_comb_pc <- tse_data_comb %>% mutate(
  pc_chg=VALUE/lag(VALUE)-1
)
```


```{r}
## join data
#infl_tse <- inner_join(infl_data_main, tse_data_main, by='date')
## better to bind rows
infl_tse <- bind_rows(infl_data_comb_pc, tse_data_comb_pc)
```

```{r}
infl_tse %>% ggplot(aes(x=date, y=pc_chg, color=item))+geom_line()
```
* Stock returns are MUCH more volatile than inflation rate!

```{r}
infl_tse_cor <- infl_tse %>% filter(!is.na(pc_chg))
infl_cor <- infl_tse_cor[infl_tse_cor$item=='All-items',"pc_chg"]
infl_cor <- infl_cor[-624,]
tse_cor <- infl_tse_cor[infl_tse_cor$item=='TSE Index close', "pc_chg"]
cor.test(tse_cor$pc_chg, infl_cor$pc_chg)
```


Probably need to look at longer time horizon to smooth out the volatility.

* compare rolling 12-mth changes

```{r}
infl_tse_wide <- infl_tse %>% select(-UOM, -pc_chg)
infl_tse_wide <- infl_tse_wide %>% pivot_wider(names_from=item, values_from=VALUE)

```

```{r}
cor.test(infl_tse_wide$`All-items`, infl_tse_wide$`TSE Index close`)
```

