---
title: 'R Data Frame vs Time Series Object: Which one when?'
author: Package Build
date: '2022-01-22'
slug: []
categories:
  - R
tags:
  - dataviz
description: 'Exploration of data frames vs time series objects in R.'
thumbnail: ''
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message=FALSE,
                      warning=FALSE,
                      fig.height=6,
                      fig.width=9)
options(scipen = 99)
options(digits=3)

library(tidyverse)
library(xts)
library(dygraphs)
library(quantmod)
```

Data frames are one of the fundamental building blocks of data analysis done in R. In some cases, though, **time series** objects offer more flexibility and other advantages when analyzing time-based data.

So when to choose time series rather data frames?

### Time Series Object

**xts** is the main R package for working with time series. **Datacamp** has an excellent [xts Cheat Sheet](http://datacamp-community-prod.s3.amazonaws.com/72771032-0653-4d87-a798-4a83879e99c8) and very helpful [supplementary info.](https://www.datacamp.com/community/blog/r-xts-cheat-sheet)  

So I'm just going to hit some highlights with examples here.

```{r}
## use quantmod pkg to get some stock prices as time series
ea <- getSymbols(Symbols='EA', auto.assign=FALSE)
head(ea)
```

Stock market prices are available in time series form via R quantmod package. As shown above, dates are row numbers rather than a date field as they would be in typical data frame. This has a variety of implications.

* types of calculations easily available
* selecting date ranges
* aggregating into periods like weekly, monthly, quarterly, annualy
* charting

```{r}
## use dygraphs pkg
#dygraph(OHLC(ea))
dygraph(ea$EA.Close)
chart_Series(ea)
```


### Limitations

* multi-dimensional data
* faceted charts