---
title: Daily Patterns in Crypto Currency Prices
author: Package Build
date: '2022-05-28'
categories:
  - crypto currency
  - R
  - analysis
tags:
  - bitcoin
  - crypto
  - dataviz
  - finance
slug: daily-patterns-in-crypto-currency-prices
draft: yes
description: Analysis of recent Bitcoin price history to see if there are predictable
  patterns in prices based on day of week that can be leveraged for trading.
---

```{r setup, include=FALSE}
## standard settings
knitr::opts_chunk$set(echo = FALSE,
                      message=FALSE,
                      warning=FALSE,
                      fig.height=3.8,
                      fig.width=7)
options(scipen = 99)
options(digits=3)

## load packages
library(quantmod)
library(tidyverse)
library(lubridate)
library(here)
library(PerformanceAnalytics)
library(plotly)
library(scales)
library(gridExtra)
library(DT)
library(quantmod)
library(dygraphs)
library(glue)

## set ggplot theme
theme_set(theme_classic())
## chart variables
fill_color <- 'steelblue3'
```

### Get Data

* Due to long-term volatility and evolution of the market focus on recent years.
* Go back roughly 2 years from today's date as starting point.
* Using BTC-CAD because...I'm Canadian.

```{r data, echo=TRUE}
symb <- c('BTC-CAD')
date_st <- '2020-05-01'
## using auto.assign = FALSE and setting object name to avoid issues with default 'BTC-CAD' name
BTC_CAD <- getSymbols(Symbols=symb, from=date_st, to=Sys.Date()-1, auto.assign = FALSE)
```

### Initial Look at Data

```{r, echo=TRUE}
dygraph(BTC_CAD[,"BTC-CAD.Close"])
```

<br />

### Focus on Recent Trading Range

```{r}
rec <- '2021-01-01/'
btc_rec <- BTC_CAD[rec]
dygraph(btc_rec[,"BTC-CAD.Close"])
```
<br />

### Add Days of Week

* Add weekdays to identify and compare prices by day of week.
* Add date of each week to identify and compare weeks.

Easiest - for me, at least - to convert time series to data frame:

```{r, echo=TRUE}
btc_rec_df <- data.frame(btc_rec)
btc_rec_df$date <- index(btc_rec)
## add days
btc_rec_df$day <- weekdays(index(btc_rec), abbreviate = TRUE)
## - set days to factors
btc_rec_df$day <- factor(btc_rec_df$day)
btc_rec_df$day <- fct_relevel(btc_rec_df$day, levels=c("Sun","Mon","Tue","Wed","Thu","Fri","Sat"))
## add weeks
btc_rec_df$week_of <- floor_date(btc_rec_df$date, unit='weeks')
```

View the structure of the data with additional components added:

```{r}
## view structure
str(btc_rec_df)
```

### Comparative views

Take a look at daily comparisons for this period:

```{r, echo=TRUE}
btc_rec_df %>% ggplot(aes(x=day, y=BTC.CAD.Close))+geom_boxplot(fill=fill_color)+
  scale_y_continuous(labels=dollar_format())+
  labs(title="Distribution of BTC-CAD Closing Price by Day of Week", y='BTC-CAD Close', x="")
```

No strong, obvious pattern over the period. Slight pattern of lower mid-week, rising to weekends. 

* *Maybe* Fri-Sat? Sun-Tue?

This data is over a period where there is a lot of variance in the price that may blur the daily patterns. Could still be that there is a consistent pattern *within* weeks.

For more granularity, let's look at week-by-week trends by day: 

```{r, echo=TRUE}
week_plot <- btc_rec_df %>% ggplot(aes(x=day, y=BTC.CAD.Close, color=as.factor(week_of), group=week_of))+geom_line()+
  scale_y_continuous(labels=dollar_format())+
  scale_x_discrete(expand=c(0,0))+
  labs(title="BTC-CAD Closing Price by Day by Week",x="",y="BTC-CAD Close")+
  theme(legend.position = 'none')
ggplotly(week_plot)
```

Certainly some weeks with upward trend through the week (most obviously near end of 2021), but not exactly a consistent pattern to rely on across this date range. 

### Specific Day of Weeks Comparisons

Look at some individual day of week comparisons

* Start with Wed - Sat

```{r}
btc_rec_d_df <- btc_rec_df %>% filter(day=='Fri' | day=='Sat')
dd_plot <- btc_rec_d_df %>% ggplot(aes(x=day, y=BTC.CAD.Close, color=as.factor(week_of), group=week_of))+geom_line()+
  scale_y_continuous(labels=dollar_format())+
  scale_x_discrete(expand=c(0,0))+
  labs(title="BTC-CAD Closing Price by Day by Week",x="",y="BTC-CAD Close")+
  theme(legend.position = 'none')
ggplotly(dd_plot)
```

Pretty hard to pick out any obvious/consistent pattern. Let's take a closer look:

* get each week % change from Wed to Sat.
* get summary stats on these changes.
* look at distribution of % change to see if any consistency.

```{r}
## may need to remove first row if starts with the day later in the week
#btc_rec_d_df <- btc_rec_d_df[-1,]

## may need to remove last row, if ends on day earlier in the week
#btc_rec_d_df <- btc_rec_d_df[-nrow(btc_rec_d_df),]
## calc % chg Wed-Sat
btc_rec_d_df <- btc_rec_d_df %>% mutate(
  wk_chg=BTC.CAD.Close/lag(BTC.CAD.Close)-1
)
## calculate some stats and make them pretty for printing
mwkchg_calc <- median(btc_rec_d_df$wk_chg, na.rm=TRUE)
mwkchg <- glue(prettyNum(mwkchg_calc*100, digits=2),"%")
awkchg_calc <- mean(btc_rec_d_df$wk_chg, na.rm=TRUE)
awkchg <- glue(prettyNum(awkchg_calc*100, digits=2),"%")
wkchg_pctl <- quantile(btc_rec_d_df$wk_chg, 0.5, na.rm=TRUE)
## set color for mean based on above/below zero
acolor <- ifelse(awkchg_calc>0,'green','red')
apos <- ifelse(awkchg_calc>0,0.03,-0.03)
mpos <- ifelse(mwkchg_calc>0,0.04,-0.04)
## histogram
btc_rec_d_df %>% ggplot(aes(x=wk_chg))+geom_histogram(fill=fill_color)+
  geom_vline(xintercept=mwkchg_calc, color='black', linetype='dashed', size=1)+
  geom_vline(xintercept=awkchg_calc, color=acolor, linetype='dashed', size=1)+
  annotate(geom='text', label=paste0("median: ",mwkchg), x=mwkchg_calc+mpos, y=16.4, color='black')+
  annotate(geom='text', label=paste0("ave: ",awkchg), x=awkchg_calc+apos, y=17.1, color=acolor)+
  labs(title="Distribution of Weekly Returns from Fri to Sat")
  
```

Basically a wash:

* median of `r mwkchg` tells us there is 50% chance of being either above or below `r mwkchg` (very close to 0) return on the week, although farther outliers on the positive side.
* Ave. return on the week *could* add up to something over time. (`r awkchg`).

Compare to Sun to Tue:

```{r}
btc_rec_d_df <- btc_rec_df %>% filter(day=='Sun' | day=='Tue')
## may need to remove last row, if ends on day earlier in the week
btc_rec_d_df <- btc_rec_d_df[-nrow(btc_rec_d_df),]
dd_plot <- btc_rec_d_df %>% ggplot(aes(x=day, y=BTC.CAD.Close, color=as.factor(week_of), group=week_of))+geom_line()+
  scale_y_continuous(labels=dollar_format())+
  scale_x_discrete(expand=c(0,0))+
  labs(title="BTC-CAD Closing Price by Day by Week",x="",y="BTC-CAD Close")+
  theme(legend.position = 'none')
ggplotly(dd_plot)
```

```{r}
## calc % chg Sun-Tue
btc_rec_d_df <- btc_rec_d_df %>% mutate(
  wk_chg=BTC.CAD.Close/lag(BTC.CAD.Close)-1
)
## calculate some stats and make them pretty for printing
mwkchg_calc <- median(btc_rec_d_df$wk_chg, na.rm=TRUE)
mwkchg <- glue(prettyNum(mwkchg_calc*100, digits=2),"%")
awkchg_calc <- mean(btc_rec_d_df$wk_chg, na.rm=TRUE)
awkchg <- glue(prettyNum(awkchg_calc*100, digits=2),"%")
wkchg_pctl <- quantile(btc_rec_d_df$wk_chg, 0.5, na.rm=TRUE)
## set color for mean based on above/below zero
acolor <- ifelse(awkchg_calc>0,'green','red')
apos <- ifelse(awkchg_calc>0,0.03,-0.03)
mpos <- ifelse(mwkchg_calc>0,0.04,-0.04)
## histogram
btc_rec_d_df %>% ggplot(aes(x=wk_chg))+geom_histogram(fill=fill_color)+
  geom_vline(xintercept=mwkchg_calc, color='black', linetype='dashed', size=1)+
  geom_vline(xintercept=awkchg_calc, color=acolor, linetype='dashed', size=1)+
  annotate(geom='text', label=paste0("median: ",mwkchg), x=mwkchg_calc+mpos, y=16.4, color='black')+
  annotate(geom='text', label=paste0("ave: ",awkchg), x=awkchg_calc+apos, y=17.1, color=acolor)+
  labs(title="Distribution of Weekly Returns from Sun to Tue")
```

